Towards Proof-Reconstruction of Problems in Classical Propositional Logic

### 2017-09-08 paper/sections/abstract.tex
[ paper ] fixing error grammars and adding notes.

\begin{abstract}
In this paper, we describe our approach to reconstruct and verify in
\Agda, a  proof assistant,  solutions obtained by \Metis, an
automatic theorem prover, to problems in classical propositional
logic. The translation method consists of justifying step-by-step
each deduction found in the prover's solution. To carry out this
plan, we emulate the inference rules of the prover, providing a
version of them in the proof assistant. Finally, we parse the output
of the prover,  constructing the respective tree associated with the
proof, ending up with an \Agda code for type-checking.

\keywords{proof reconstruction, \Agda, automatic theorem prover, classical propositional logic, \Metis.}
\end{abstract}

>>>>

\begin{abstract}
In this paper, we describe our approach to reconstruct and verify in
\Agda, a  proof assistant for type theory, solutions obtained by \Metis, an
automatic theorem prover, to problems in classical propositional
logic. The translation method consists of justifying step-by-step
each deduction found in the prover's solution. To carry out this
plan, we emulate the inference rules of the prover, providing a
version of them in the proof assistant. Finally, we parse the output
of the prover, build and analize the respective tree associated with the
proof, for ending up with an \Agda code of the proof for type-checking.

\keywords{proof reconstruction
\unsure{proof reconstruction or proof-reconstruction},
\Agda, automatic theorem prover, classical propositional logic, \Metis.}

\end{abstract}

### 2017-09-08 paper/sections/acknowledgements.tex
[ paper ] fixing error grammars and adding notes.

We thank Universidad EAFIT of Medell\'in, Colombia for funding
support. This project was a product of the Logic and Computation
Research Group at the Universidad EAFIT.

>>>>

We thank Universidad EAFIT in Medell\'in, Colombia for funding
support. This project was a product of the Logic and Computation
Research Group at Universidad EAFIT.

### 2017-09-08 paper/sections/conclusions.tex
[ paper ] fixing error grammars and adding notes.

Amazing conclusions.
...
\cite{}

\subsubsection*{Future Work.}
\label{ssec:future-work}

>>>>

Besides the logic and its complexity, the proof reconstruction task is
far from trivial. In this document, we attempt to reconstruct proofs
of propositional logic without equality deliverd by \Metis successfully.

From our experience building a proof reconstruction tool, depending on
the systems involved in thetranslation, some difficulties can arise.
We listed some of them in section \ref{sec:metis-language-and-proofs}
and in the section \ref{sec:proof-reconstruction}.

For instance, just showing equality between Normal Forms for the
same formula in the output of the \canonicalize inference and a
common implementation for this kind of conversion; we observed that
\Metis performs reordering for each inner expression mainly in the
presence of negative literals. Another aspect to take into account
is the \Metis' treatment for conjunctions and disjunction;
these connectives were treated as right associative and binary
operators in the \TSTP format but that assumptons is not a standard.
We had to update the \TSTP parser module introduced in
\cite{Gomez-Londono2015} to include features like these.

However, one of our biggest obstacles
\improvement{Remove the use of the \emph{biggest obstacules}.}
was dealing with
the Equality between formulas since, in our approach, we only work
with the syntactic treatment of the logic without considering the
semantics.
One of the main reasons for such a restriction relay on our plans to
support further first-order logic, where a semantic treatment is not
viable.\unsure{Please confirm this.}

Nevertheless, the proof reconstruction is certainly possible, and
the verification of solutions of \CPL problems can be carried out
successful in \Agda. Our tool is able to translate \Metis
derivations in \TSTP format to deliver \Agda code of the proof,
\change{Better uses \emph{proof-terms}?}
and this proof can be type-checked by \Agda.

The source code of our tool can be extended to support
other \ATPs for \CPL like \name{E}. This work can be carried out
may following the description in Kanso' Ph.D. Thesis~\cite{Kanso2012}
for the inference rules of \name{E} or \name{Z3}.

Trying to justify a proof-object of a theorem prover has a
significant impact, may increase the trustworthiness of the prover
proving its correctness, or contrary, revealing important issues to
improve the system. In both cases, there is a benefit for both
sides, the user of this technology, and their developers.

For instance, while writing this document, we found a soundness bug,
a bug in the printing out of the \TSTP derivations, and a bug in the
parsing for \TPTP problem in \Metis. Fortunately, all these bugs
were fixed quickly by Hurd\footnote{Issues No. 4, No. 2, and the
commit \name{8a3f11e} from the \Metis' official repository
\url{https://github.com/gilith/metis}}, and a new version of \Metis
was released (version 2.3, release 20170810).

% \subsubsection*{Future Work.}
% \label{ssec:future-work}

### 2017-09-08 paper/sections/introduction.tex
[ paper ] fixing error grammars and adding notes.

A proof reconstruction tool provides such an integration translating
the reply delivered by the prover into the formalism of the proof
assistant. Because the formalism of the source (the evidence
generated by the \ATP) is not necessarily the same logic in the
target, the reconstruction turns out in a ``reverse engineering''
task. Then, reconstructing a proof involves an in-depth
understanding of the algorithms in the \ATP and the specification
logic in the \ITP. What we need from the \ATP tools is a
proof-object in a consistent format to work with, that is, a full
script describing step-by-step with exhaustive details and without
ambiguities, the table of the derivation to get the actual proof.
For problems in classical propositional logic (henceforth \CPL),
from a list of at least forty\footnote{\ATPs available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs (e.g., \name{CVC4}
\cite{Barrett2011}, \name{SPASS}, and \name{Waldmeister}
\cite{hillenbrand1997}) and a little bit less replied with a proof
in a file format like \TSTP~\cite{sutcliffe2004tstp} (e.g.,
\name{E}, \Metis, \name{Vampire}, and \name{Z3}), \len{LFSC}
\cite{Stump2008}, or the \len{SMT-LIB}~\cite{Bohme2011} format.

Many approaches have been proposed and some tools have been
implemented for proof reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013}, \cite{Bohme2011}, \cite{Fleury2014} and
\cite{Kanso2012}). % We mention some tools in the following.

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows. In section
\ref{sec:metis-language-and-proofs}, we briefly introduce the \Metis
prover. In section \ref{sec:proof-reconstruction}; we present our
approach to reconstruct proofs deliver by \Metis in \Agda. We list
related work in proof reconstruction in section
\ref{sec:related-work}. In section~\ref{sec:conclusions}, we discuss
some limitations, advantages, and disadvantages with other similar
proof reconstruction tools and conclusions for ending up with the
future work.

>>>>

A proof reconstruction tool provides such an integration translating
the reply delivered by the prover into the formalism of the proof
assistant. Because the formalism of the source (the evidence
generated by the \ATP) is not necessarily the same logic in the
target, the reconstruction turns out in a ``reverse engineering''
task. Then, reconstructing a proof involves an in-depth
understanding of the algorithms in the \ATP and the specification
logic in the \ITP. To begin with the proof reconstruction, it is
necessary to have a proof-object from the \ATP in a consistent
format to work with, that is, a full script describing step-by-step
with exhaustive details and without ambiguities, the derivation to
get the actual proof.
For problems in classical propositional logic (henceforth \CPL),
from a list of at least forty\footnote{\ATPs available from the web
service \name{SystemOnTPTP} of the TPTP World.} \ATPs, just a few
provers were able to deliver proofs (e.g., \name{CVC4}
\cite{Barrett2011}, \name{SPASS}, and \name{Waldmeister}
\cite{hillenbrand1997}) and a little bit less replied with a proof
in a file format like \TSTP~\cite{sutcliffe2004tstp} (e.g.,
\name{E}, \Metis, \name{Vampire}, and \name{Z3}), \len{LFSC}
\cite{Stump2008}, or the \len{SMT-LIB}~\cite{Bohme2011} format.

Many approaches have been proposed and some tools have been
implemented for proof reconstruction in the last decades. These
programs are relevant not only because it helps to spread their
usage but they also increase the confidence of their users about
their algorithms and their correctness (see, for example, bugs in
\ATPs~\cite{Keller2013}, \cite{Bohme2011}, \cite{Fleury2014} and
\cite{Kanso2012}). We mention some of these tools in the related work,
section \ref{sec:related-work}.

In this paper, we describe the integration of \Metis prover with the
proof assistant \Agda by parsing \TSTP derivations to generate \Agda
proof-terms. We structure the paper as follows. In section
\ref{sec:metis-language-and-proofs}, we briefly introduce the \Metis
prover. In section \ref{sec:proof-reconstruction}, we present our
approach to reconstruct proofs deliver by \Metis in type theory.
In section~\ref{sec:conclusions}, we discuss
some limitations, advantages, and disadvantages with other similar
proof reconstruction tools for ending up with some conclusions and the
future work. Finally, we list some related works in proof reconstruction
in section \ref{sec:related-work}.

### 2017-09-08 paper/sections/metis-language-and-proofs.tex
[ paper ] fixing error grammars and adding notes.

\Metis is an automatic theorem prover written in \len{Standard ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. Exists tools that perform proof reconstruction with \Metis,
some of these systems are the following.

@@

In this paper, we ported a subset of \Metis' inference rules to
\Agda, the propositional fragmented, to allow us to justify
step-by-step the proofs delivered in \TSTP format generated
by \Metis.

@@

For instance, we express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \vee \neg q)$
in \TPTP syntax as follows, declaring an \tptpaxiom $p$, and the
expected conclusion, the  \tptpconjecture
\verb!~ ((p & ~ p) | (q & ~ q))!.

@@

These proofs are directed acyclic graphs, trees of refutations. Each
node stands for an application of an inference rule and the leaves
in the tree represent formulas in the given problem. Each node is
labeled with a name of the inference rule (e.g., \canonicalize in
Fig.~\ref{fig:metis-example}). Each edge links a premise with one
conclusion. The proof graphs have at their root the conclusion $\bot$
since \Metis delivers proof by refutation.

@@

For instance, for the propositional logic fragment without equality,
only three rules are valid from the \Metis' Logic Kernel:
\emph{assume}, \emph{axiom}, and \emph{resolve}. Nevertheless, as we
observed in \TSTP derivations, for such a propositional fragment,
\Metis uses in his deductions steps, six rules where only the
\emph{resolve} inference persists from the original list. In the
case of \resolve, it differs from its original version since, in
\TSTP, this rules come with an argument, a literal. The other
undocumented rules found in \TSTP derivations are: \canonicalize,
\clausify, \conjunct, \negate, \simplify, and \strip.

This lack of the \Metis' documentation was an important matter for
us in the proof reconstruction and certainty for other implementors
(see a comment related on the presenting of \name{Sledgehammer}
in \cite{paulson2007source} and \cite{Farber2015}). We faced
the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify. For example, we believe that with these last
inferences, \Metis summarized many essential steps in the conversion
or deduction from one formula or a list of formulas to another as we see in the following example.
\improvement{Add an example showing a deduction where metis do a lot 
between two steps}
That is inconvenient since, our proof reconstruction approach needs
take into account any assumption or consideration as ordering for
literals, application of simplification theorems, reductions, among
other valid transformations.

@@

In this part, our main purpose consists of giving a response to
these rules looking at the \Metis' source code while inspecting a
representative set of solutions of \CPL problems
\cite{Prieto-Cubides2017} generated by \Metis.
This is an attempt to fulfill the \Metis's documentation,
describing our understanding of each rule, to later on in section
\ref{sec:proof-reconstruction}, formalize them effectively in \Agda.

@@

\textit{Normalization.} \Metis uses the \canonicalize rule to
normalize a propositional formula that comes from an axiom or a
definition. It often converts each expression in one of its normal
form, the conjunctive normal form, the negative normal form or the
disjunctive normal form. Posterior, this rule simplifies the formula
with some definitions, applying recursively on the structure's
formula conjunctions and disjunctions simplifications to remove
tautologies (see Fig.~\ref{fig:conjunctive-disjunctive-simpl} with a
list of these theorems. We had assumed the commutative property for
conjunction and disjunction connectives).

@@

\textit{Resolution.} The \resolve rule is the resolution theorem
showed in Fig.~\ref{fig:metis-inferences}. Its applications needs
the positive literal for resolution and two derivations to deduce
the \emph{resolvent}. The positive literal $p$ must occur in the
formula from the first derivation, and the negative literal must
occur in the formula from the second derivation (this follows the
same pattern in \Metis' logic core in Fig.~\ref{fig:metis-inferences}
for the resolve inference).

@@

\textit{Splitting a conjunct.} The \conjunct rule extracts from a
conjunction one of its conjuncts; it is a generalization of the
projection rules for the conjunction connective as we can see in the
following \TSTP derivation.

@@

\textit{Simplification.}
The \simplify rule is inference that performs simplification of 
definitions or tautologies over a list of derivations. This rule 
transverses a list of formulas by applying different theorems (e.g.,
\emph{modus pones}, \emph{modus tollens}, or \emph{disjunctive
syllogism}) including the list of theorems used by \canonicalize in Fig.
\ref{fig:conjunctive-disjunctive-simpl} and the resolution theorem of
\resolve rule. Moreover, many things happen inside this \Metis'procedure, 
and it is somewhat complicated to grasp it completely.
\end{document}

>>>>

\Metis is an automatic theorem prover written in \len{Standard ML}
for first-order logic with equality developed by John Hurd
\cite{hurd2003first}. This prover is suitable for proof
reconstruction since it provides well-documented proof-objects to
justify its deduction steps from the basis of only six inference
rules. There are already some tools that perform proof reconstruction
with \Metis, some of these systems are the following.

@@

In this paper, we ported a subset of the \Metis' inference rules to
\Agda, the propositional fragmented without equality.
That allows us to justify step-by-step the proofs delivered in
\TSTP format generated by the prover \Metis.

@@

For instance, let us express the problem
$p\, \vdash \neg (p \wedge \neg p) \vee (q \vee \neg q)$
in \TPTP syntax as follows, declaring an \tptpaxiom $p$, and the
expected conclusion, the  \tptpconjecture
\verb!~ ((p & ~ p) | (q & ~ q))!.

@@

These proofs are directed acyclic graphs, trees of refutations. Each
node stands for an application of an inference rule and the leaves
in the tree represent formulas in the given problem. Each node is
labeled with a name of the inference rule (e.g., \canonicalize in
Fig.~\ref{fig:metis-example}). Each edge links a premise with one
conclusion. The proof graphs have at their root the conclusion
$\bot$, since \Metis delivers proofs by refutation.

@@

For instance, for the propositional logic fragment without equality,
only three rules are valid from the \Metis' Logic Kernel:
\emph{assume}, \emph{axiom}, and \emph{resolve}. Nevertheless, as we
observed in \TSTP derivations, for such a propositional fragment,
\Metis uses in his deductions steps, six rules where only the
\emph{resolve} inference persists from the original list. In the
case of \resolve, it differs from its original version since, in
\TSTP, this rule has by definition an argument, a literal. The other
undocumented rules found in \TSTP derivations are: \canonicalize,
\clausify, \conjunct, \negate, \simplify, and \strip.

This lack of the \Metis' documentation is an important matter for
us in the proof reconstruction and certainty for other implementors
(see a comment related on the presenting of \name{Sledgehammer}
in \cite{paulson2007source} and \cite{Farber2015}). We faced
the ambiguity and lack of clearness of some rules like \canonicalize
or \simplify.
For example, we believe by applying these last
inferences, \Metis summarized many essential steps to deduce a
formula from a list a list of formulas as we see in the following
example.
\improvement{Add an example here to show a deduction where \Metis do
a lot between two steps.}
That inference' behavior has become a drawback since, our approach
to reconstruct proofs needs take into account any assumption or
consideration as ordering for literals, application of
simplification theorems, reductions, among other valid
transformations.

@@

In this part, our main purpose consists of giving a response to
these rules looking at the \Metis' source code while inspecting a
representative set of solutions of \CPL problems
\cite{Prieto-Cubides2017} generated by \Metis.
This is an attempt to fulfill the \Metis's documentation, by
describing our each rule from our perspective. Later on, in section
\ref{sec:proof-reconstruction}, we present a formal description of
such rules.

@@

\textit{Normalization.} \Metis uses the \canonicalize rule to
normalize a propositional formula that comes from an axiom or a
definition. It often converts each expression in one of its normal
form: the conjunctive normal form, the negative normal form or the
disjunctive normal form. Next, this rule simplifies the formula
with some definitions, applying recursively on the structure's
formula conjunctions and disjunctions simplifications to remove
tautologies (see Fig.~\ref{fig:conjunctive-disjunctive-simpl} for a
list of these theorems. We assume the commutative property for the
conjunction and disjunction connectives).

@@

\textit{Resolution.} The \resolve rule is the resolution theorem
showed in Fig.~\ref{fig:metis-inferences}. For its application, we
need the positive literal for resolution and two derivations to
deduce the \emph{resolvent}. The positive literal $p$ must occur in
the formula from the first derivation, and the negative literal must
occur in the formula from the second derivation (this follows the
same pattern in \Metis' logic core in Fig.~\ref{fig:metis-inferences}
for the resolve inference).

@@

\textit{Splitting a conjunct.} The \conjunct rule extracts from a
conjunction one of its conjuncts; it is a generalization of the
projection rules for the conjunction connective as the
following \TSTP derivation shows.

@@

\textit{Simplification.}
The \simplify rule is inference that performs simplification of
definitions or tautologies over a list of derivations. This rule
transverses a list of formulas by applying different theorems (e.g.,
\emph{modus pones}, \emph{modus tollens}, or \emph{disjunctive
syllogism}) including the list of theorems used by \canonicalize in
Fig. \ref{fig:conjunctive-disjunctive-simpl} and the resolution
theorem of \resolve rule.

Moreover, we are not completely sure about the exact behavior for
all \Metis' inference rules. It is somewhat complicated to grasp it
completely since the source code is around the 10.000 lines of code.

\end{document}
